Theory of phonon polaritons in polar dielectrics

### 2018-08-02 paper.tex
update manuscript
made edits discussed yesterday

\begin{abstract}
The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.  Largely, the manifestation of this idea has been plasmonics, which have been demonstrated to confine electromagnetic fields on the scale of nanometers. A more recent development is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride. These modes pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band. To further advance this promising field, it is of great interest to compute the optical response of polar dielectrics from first principles. In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the nonlocal dielectric function. While in some limits, our results qualitatively match the famous Lorentz oscillator model, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  phonon band-structure, and the phonon polarizations.  Moreover, our results allow us to point to a regime where quantum nonlocal effects may correct the phonon polariton dispersion, which may be relevant in experiments which confine electromagnetic fields to the scale of 1 nm. Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of ab-initio-derived parameters of the underlying lattice. We derive the electrodynamics of 2D phonon polaritons, finding their fields and dispersion. Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide, and show that they agree well with experimental data.
\end{abstract}

@@

There are nevertheless some outstanding questions regarding the behavior of phonon polaritons. Here, we present a few. 
\begin{enumerate}
\item{Can the very successful Lorentz oscillator model ever break down? One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion. Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick).}
\item{What are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators? Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene?}
\item{In general, are the properties of optical phonons inherited by the phonon polariton? What is the influence of concepts such as interface phonons and phonon topology on the polariton? The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic materials.}
\end{enumerate}

Indeed, the way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings. Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.

@@

\textcolor{blue}{Pri: one place we can get criticized here by an experimentalist (or theorist) is that at least for the 3D spatially-local case, there's this thing that Caldwell called the TO-LO formalism in his proposal. It's not actually a formalism... it's just an expression for the dielectric function in terms of the high-frequency electronic part, and the various TO and LO phonon frequencies, as well as the damping. It features some fairly simple-to-get experimental parameters (though actually I don't know how they get the LO frequency since only TO comes up in Raman I think), and reproduces the data generally pretty well. I think actually this is what they did in that arXiv paper that worried us about a month ago. Anyway, when we do nonlocal and reduced-dimensional stuff, I think this is where we can really be different. In any event, it's good to anticipate this criticism in advance so we can potentially get in front of it. What do you think?}

\textcolor{blue}{To all: I'm new to phonons and am likely making some non-standard notational choices? If so, feel free to point out where and I'll try to standardize it.} We start by reviewing the theory of phonon polaritons in a 3D bulk, developing the theory in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties. Here, the matter subsystem which couples to light is the ionic lattice of the crystal. Thus, we develop the linear response theory of an ionic lattice to an electromagnetic field. The interaction Hamiltonian of the lattice with the electromagnetic field is taken as
\begin{equation}
H_{int} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
\end{equation}
where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice. We write the polarization as a sum of contributions from all unit cells in the system. Namely
\begin{equation}
\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
\end{equation}
where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$. In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation. This said, we operate in the harmonic regime.  The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa} \mathbf{b}_{\kappa} \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$. Note that the Born effective charges are both tensorial and site-dependent. \textcolor{blue}{I still want to give a decent intuition for these Born effective charges as I find them a bit mysterious and so will a reader who is from photonics.} This displacement can be expanded in phonon modes as
\begin{equation}
\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
\end{equation}
where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation (1). We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.

@@

Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. 
\textcolor{blue}{Should this part below be moved to the outlook or does it make sense to be here instead?} Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

@@

\textcolor{red}{ Should we mention here that while we take the dissipation rate and bandgap from experiment, that these quantities can be obtained ab initio as well? We could cite packages that would allow someone to do this for a material where properties aren't well known, just to keep ourselves in the "first principles" designation. Maybe a note in the supplement or a footnote with this to prevent "semi-empirical" finger pointing?}

\textcolor{blue}{I included a sentence addressing this important point after the sentence where we say that we take the lifetimes from experiment. }

@@

\textcolor{blue}{Pri: blue text below means this particular suggestion is heavily inspired from what Caldwell sent us.} In summary, we have provided a theoretical framework based on linear response theory to calculate the phonon contribution to the dielectric function from first principles. Our framework is rather versatile, allowing us to: use first principles calculations to get the dielectric function, predict how nonlocality enters the dielectric function, and predict the effect of reduced dimensionality. A particularly interesting case to consider in future work would be to find a system where the optical phonons are drastically different from 3D to 2D. Perhaps it is possible that there are some materials in which the 2D optical phonons experience lower losses due to a reduced scattering phase space. The formalism we provide here may also be extended to understand phonon-polaritons in other more atypical reduced-dimensional settings, such as zero-dimensional settings in single emitters, i.e., "molecular phonon polaritons", in analogy to recent work on "molecular plasmons" \cite{manjavacas2013tunable,lauchner2015molecular}. Notably, we go beyond oversimplifications of the Lorentz oscillator model in which the Born charges are treated as a single scalar quantity and can treat the influence and interplay of many phonon modes that contribute to the dielectric function. We corroborated our approach through density functional theoretic calculations.

In the long run, besides furnishing an ab-initio description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here. For example, we may consider situations such as \textcolor{blue}{the phonon polaritons associated with optical phonons at the interface of two materials. Another situation of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.} In that case, it would be of interest to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.

>>>>

\begin{abstract}
The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.  An important recent development along this paradigm is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride, which pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band. To further advance this promising field, it is of great interest to predict the optical response of polar dielectrics from first principles. In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the dielectric function. For a 3D bulk in the spatially local regime, our results qualitatively match the famous Lorentz oscillator model. However, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  and the phonon-bandstructure.  Moreover, our results allow us to point to a regime where quantum nonlocal effects may correct the phonon polariton dispersion, which may be relevant in experiments which confine electromagnetic fields to the scale of 1 nm. Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same ab-initio-derived parameters of the underlying lattice. We derive the electrodynamics of 2D phonon polaritons, finding their fields and dispersion. Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide, and show that they agree well with experimental data. Our results provide a stepping stone for eventually considering from first principles the effects of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
\end{abstract}

@@

There are nevertheless some outstanding questions regarding the behavior of phonon polaritons. Here, we present a few. First, can the very successful Lorentz oscillator model ever break down? One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion. Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick). Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators? Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene? In general, are the properties of optical phonons inherited by the phonon polariton? Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton? Examples of non-trivial spatial shaping include chiral phonons, interface phonons, and mechanical heterostructurs. The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic nano-materials.

The way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings. Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.

@@

We start by developing the theory of phonon polaritons in a 3D bulk in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties. Here, the matter subsystem which couples to light is the ionic lattice of the crystal. Thus, we develop the linear response theory of an ionic lattice to an electromagnetic field. The interaction Hamiltonian of the lattice with the electromagnetic field is taken as
\begin{equation}
H_{int} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
\end{equation}
where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice. We write the polarization as a sum of contributions from all unit cells in the system. Namely
\begin{equation}
\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
\end{equation}
where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$. In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation. This said, we operate in the harmonic regime.  The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa} \mathbf{b}_{\kappa} \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$. Note that the Born effective charges are both tensorial and site-dependent.  This displacement can be expanded in phonon modes as
\begin{equation}
\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
\end{equation}
where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation (1). We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.

@@

Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

@@

@@

In summary, we have provided a theoretical framework based on linear response theory to calculate the phonon contribution to the dielectric function from first principles. Our framework is rather versatile, allowing us to: use first principles calculations to get the dielectric function, predict how nonlocality enters the dielectric function, and predict the effect of reduced dimensionality. A particularly interesting case to consider in future work would be to find a system where the optical phonons are drastically different from 3D to 2D. Perhaps it is possible that there are some materials in which the 2D optical phonons experience lower losses due to a reduced scattering phase space. The formalism we provide here may also be extended to understand phonon-polaritons in other more atypical reduced-dimensional settings, such as zero-dimensional settings in single emitters, i.e., "molecular phonon polaritons", in analogy to recent work on "molecular plasmons" \cite{manjavacas2013tunable,lauchner2015molecular}. Notably, we go beyond oversimplifications of the Lorentz oscillator model in which the Born charges are treated as a single scalar quantity and can treat the influence and interplay of many phonon modes that contribute to the dielectric function. We corroborated our approach through density functional theoretic calculations.

In the long run, besides furnishing an ab-initio description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here. For example, we may consider situations such as the phonon polaritons associated with optical phonons at the interface of two materials. Another situation of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other. In that case, it would be of interest to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
